---
title: "A Comparative Analysis of Student GPA and Career Readiness in 2026"
subtitle: "An Analysis of Usage Patterns and Academic Performance"
author: "Rico Araojo"
date: today
format: 
  html:
    toc: true            
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: true
    embed-resources: true
    execute:
      echo: false
      message: false
      warning: false
---

```{r}
# Import libraries
# Copy your key lines from the .r file here
library(readxl)   # For importing Excel files
library(tidyverse) # For cleaning, transformation, and visualization
library(janitor)   # For quick data cleaning
library(lubridate) # For handling date formats
library(DT)        # Data Table packageS
library(tibble)
library(kableExtra)
library(knitr)
library(modelsummary)

# Import the dataset
file_path <- "AI_Impact_Student_Life_2026.csv"
df <- read_csv(file_path)
```

## Objective
To evaluate whether the widespread adoption of LLMs in higher education functions primarily as a tool for academic stabilization and time management, and to determine if this technical integration translates into measurable professional confidence or if such confidence remains dependent on human-centric, non-computable traits.

## Summary
The "How AI is Changing Life | AI and Student Life 2026" dataset, sourced from Kaggle, provides a simulated exploration of the impact of Artificial Intelligence on higher education. Artificially generated by Sohaib Malik for educational purposes, this dataset serves as a proxy for analyzing modern academic trends, AI tool adoption, and student performance metrics in a mid-2020s context.
```{r} 
metadata <- tribble(
  ~Fields, ~Definition, ~`Data Type/Format`,
  "Student_ID",  "A unique identifier for each student.",  "character",
  "Primary_AI_Tool",  "The main AI platform the student uses for academic work.",  "character",
  "GPA_Baseline",  "The student's Grade Point Average before integrating AI into their study routine.",  "numeric",
  "AI_Ethics_Concern",  "Level of concern regarding academic integrity and original work (Low/Medium/High).",  "character/factor",
  "Age",  "Age of the student, ranging from 18 to 25 years.",  "numeric",
  "Task_Frequency_Daily",  "Estimated number of times a student interacts with AI tools per day.",  "numeric",
  "GPA_Post_AI",  "The student's current Grade Point Average after adopting AI tools.",  "numeric",
  "Career_Confidence_Score",  "Self-reported score (1-10) on how prepared the student feels for an AI-driven job market.",  "numeric/factor",
  "Major",  "The student's field of study (e.g., Software Engineering, Business, Arts).",  "character",
  "Main_Usage_Case",  "The primary way AI is utilized (e.g., Debugging, Research, Exam Prep).",  "character",
  "Time_Saved_Hours_Weekly",  "Estimated hours saved per week by using AI for manual or repetitive tasks.",  "numercic"
)
metadata %>%
  kable(
    format = "html", 
    # We wrap the text in a span tag to force the alignment
    caption = "<span style='display: block; text-align: left; font-weight: bold; color: #333;'>Variable Definitions and Structural Metadata</span>"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  ) %>%
  column_spec(2, width = "25em") %>%
  column_spec(3, italic = TRUE)
```

## Key Insights
1. **AI as a Productivity "Force Multiplier"**  
The most significant impact of AI on students is time reclamation. With a median of 9 hours saved per week, AI is effectively acting as a productivity tool that allows students to either deepen their learning or manage academic burnout. This high utility explains the competitive market, where adoption is evenly spread across tools like Claude, Gemini, and Copilot.  

2. **Academic Growth - Subtle but Present**  
AI engagement shows a weak positive relationship with academic performance. While it doesn't provide a "dramatic boost," there is a clear trend of students moving away from lower GPA tiers (2.5–2.8) toward the middle and higher distributions. This suggests AI may be helping lower-performing students bridge the gap to academic stability.  

3. **The Complexity of Career Confidence**  
Professional confidence is difficult to predict through academic background alone. While Business and Data Science majors show a small, statistically significant correlation with higher confidence, the extremely low Adjusted $R^2$ (0.004) indicates that confidence is driven by factors far beyond a student’s major or AI usage.  

4. **Diverse Ethical Sentiments**  
There is no "one-size-fits-all" attitude toward AI ethics. User concern is highly dependent on the tool being used:
    * Perplexity users are the most ethically critical.
    * GitHub Copilot and Claude 3.5 users report the lowest levels of concern.

## Exploratory Data Analysis

#### Regression Model
The **regression analysis** examines factors influencing a Career Confidence Score based on a sample of 1,500 observations. The model looks at specific majors, task frequency, and time saved to see what correlates with professional confidence.  

The model suggests that academic background specifically Business and Data Science plays a small but measurable role in how confident individuals feel about their careers. However, because the Adjusted $R^2$ is 0.004, this model is a very weak predictor of overall confidence. While the intercepts and certain majors are technically significant, they represent only a tiny piece of the overall "confidence puzzle".
```{r}
model <- lm(Career_Confidence_Score ~ Task_Frequency_Daily + Major + Time_Saved_Hours_Weekly, data = df)
modelsummary(model,
             stars = TRUE,
             fmt = 3,
             # Term on rows, Model & Statistic (Estimate/P-value) on columns
             shape = term ~ model + statistic, 
             statistic = "p.value", # Tells it to use p-values for the 2nd column
             coef_rename = c("Task_Frequency_Daily" = "Daily Task Frequency",
                             "MajorData Science" = "Data Science Major",
                             "MajorBusiness Administration" = "Business Administration",
                             "MajorFine Arts" = "Fine Arts",
                             "MajorModern History" = "Modern History",
                             "MajorSoftware Engineering" = "Software Engineering",
                             "Time_Saved_Hours_Weekly" = "Weekly Hours Saved"),
             gof_omit = 'AIC|BIC|Log.Lik|RMSE',
             output = "kableExtra") %>%
  add_header_above(c(" " = 1, "Career Confidence Score" = 2)) %>% # Span title over 2 columns
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F)
```  

#### Shift in Academic Distribution
There is a visible reduction in the frequency of scores between the 2.5 and 2.8 range in the Post-AI group compared to the Baseline. This suggests that students who previously occupied the lower end of the spectrum have moved toward the middle or higher tiers of the distribution.
```{r}
baseline_post_AI_histogram <- pivot_longer(df, 
                          cols = c(GPA_Baseline, GPA_Post_AI), 
                          names_to = "Group", 
                          values_to = "GPA")

ggplot(baseline_post_AI_histogram, aes(x = GPA, fill = Group)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity", color = "white") +
  # Add vertical lines for the means
  geom_vline(xintercept = 3.257, linetype = "dashed", color = "red", linewidth = 1) +
  geom_vline(xintercept = 3.345, linetype = "dashed", color = "blue", linewidth = 1) +
  # Styling
  scale_fill_manual(values = c("GPA_Baseline" = "gray", "GPA_Post_AI" = "skyblue")) +
  labs( subtitle = "Red dashed: 3.257 (Baseline) | Blue dashed: 3.345 (Post-AI)",
       x = "GPA Score", y = "Frequency") +
  theme_minimal()
```

#### Impact of AI Engagement Frequency on Academic Growth
The graph indicates a negligible to weak positive relationship between the frequency of AI engagement and academic growth. While AI usage is generally associated with a slight improvement in GPA (as the trend line stays above zero), increasing the daily frequency of use does not appear to provide a dramatic boost in performance.
```{r}
df <- df %>% mutate(GPA_Change = GPA_Post_AI - GPA_Baseline) # add new column to the existing dataframe
ggplot(df, aes(x = Task_Frequency_Daily, y = GPA_Change)) +
  geom_jitter(alpha = 0.4) +
  geom_smooth(method = "lm", color = "red") +
  labs(x = "Daily task frequency", y = "Difference of GPA After and Before AI")
```  

#### Ethical Concern Density across Primary AI Tool Ecosystems
The heatmap reveals a clear divergence in how users of different AI platforms perceive ethical risks. While GitHub Copilot and Claude 3.5 anchor the "Low" concern spectrum with high-density clusters, Perplexity exhibits a notably higher concentration of users with "High" ethical concerns.
```{r}
df$AI_Ethics_Concern <- factor(df$AI_Ethics_Concern, 
                               levels = c("Low", "Medium", "High"), 
                               ordered = TRUE)
heatmap_data <- df %>%
  group_by(Primary_AI_Tool, AI_Ethics_Concern) %>%
  tally()

ggplot(heatmap_data, aes(x = Primary_AI_Tool, y = AI_Ethics_Concern, fill = n)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "darkred") +
  labs(x = "AI Tools", y = "Ethics Concern") +
  theme_minimal()
```

#### Distribution of AI Tools
The chart displays total volume while simultaneously breaking down the internal composition of each platform's user base. With total student counts for each tool hovering between 280 and 310, the data suggests a highly competitive landscape in which no single AI tool has yet monopolized the student market.
```{r}
ggplot(df, aes(x = Primary_AI_Tool, fill = Main_Usage_Case)) +
  geom_bar(width = 0.85, color = "white", linewidth = 0.2) +
  scale_fill_viridis_d(option = "plasma", direction = -1) +
  theme_minimal(base_size = 14) +
  labs(
    y = "Total Student Count",
    x = "AI Tool Platform", # Re-added since you want to style it
    fill = NULL             # This removes the "Usage Case" title from the legend
  ) +
  theme(
    legend.position = "bottom",
    panel.grid.major.x = element_blank(),
    # Styling the X-axis title: grey, smaller (size 10), and plain (not bold)
    axis.title.x = element_text(color = "grey50", size = 10, face = "plain"),
    axis.title.y = element_text(color = "grey50", size = 10, face = "plain"),
    legend.text = element_text(size = 8),      # Makes the text smaller
    legend.key.size = unit(0.3, "cm"),         # Makes the colored boxes smaller
    legend.spacing.x = unit(0.1, "cm"),      # Tightens space between box and text
    # Adding some spacing to the legend to prevent crowding
    legend.margin = margin(t = -10) 
  ) +
  # Use nrow = 2 if the text keeps cutting off, otherwise keep 1
  guides(fill = guide_legend(nrow = 1, byrow = TRUE))

```  

#### Analysis of Weekly Time Saved
The analysis confirms that AI tools provide a substantial productivity boost for the student population. With a median saving of 9 hours per week, AI is effectively functioning as a "force multiplier," allowing students to either reinvest that time into deeper learning or alleviate the pressures of a heavy academic workload. This significant time reclamation likely explains the high adoption rates seen across various AI platforms.
```{r}
ggplot(df, aes(y = Time_Saved_Hours_Weekly)) +
  geom_boxplot(fill = "lightgreen", outlier.colour = "red", outlier.shape = 16, outlier.size = 3) +
  labs(y = "Hours Saved Weekly",
       x = "") +
  theme_minimal() +
  scale_x_discrete(breaks = NULL) # Removes the x-axis label for a single-variable boxplot
```  

#### Distribution of Task Frequency
The chart displays a near-uniform distribution with slight fluctuations, this data shows that users are spread relatively evenly across all levels of task frequency from 1 to 10 tasks per day.
```{r}
ggplot(df, aes(x = Task_Frequency_Daily)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "white") +
  labs(x = "Tasks per Day",
       y = "Number of Users") +
  theme_minimal()
```  

## Recommendations
To capitalize on AI’s role as a productivity force multiplier, educational institutions should pivot from policing usage toward integrating AI-literacy into the core curriculum, specifically targeting students in the lower GPA tiers where the tool acts as a critical equalizer.  

Given that career confidence is largely decoupled from AI proficiency and academic major, schools should also implement holistic career development programs that focus on human-centric soft skills—such as leadership, ethical reasoning, and adaptability—which AI cannot replicate.  

Finally, developers and educators must address the "ethical divide" by fostering transparent dialogues about data provenance and algorithmic bias, ensuring that the critical mindset seen in Perplexity users becomes a standard competency across all platforms to prevent passive or unethical over-reliance.

## Conclusions
The integration of AI into the academic sphere marks a shift toward a more efficient, though ethically complex, educational landscape. While the study initially sought to determine if AI drives academic excellence, the evidence suggests its role is primarily corrective—acting as a safety net that stabilizes lower GPA distributions—rather than an accelerative force for high-tier performance.

Furthermore, the data reveals that professional self-assurance is largely decoupled from technical AI proficiency. Instead, career confidence appears to be driven by "non-computable" traits—such as leadership and adaptability—which technical integration alone cannot replicate.

Ultimately, the future of AI in education will not be defined by tool adoption, but by the user’s ability to balance productivity gains with a rigorous commitment to ethical integrity and the development of these essential human-centric skills.

## Disclaimer
This report was developed for educational purposes only. The analysis utilizes an artificially generated dataset sourced from Kaggle to simulate real-world exploratory data analysis (EDA) and professional documentation practices. All statistical modeling, including the regression analysis used to evaluate variable correlations, was performed using the R programming language. This project is intended to demonstrate data science workflows and technical proficiency in statistical computing.











